Towards a classification framework for social machines

### 2013-02-26 latex/paper_new.tex
British English

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{Workshop on the theory and practice of social machines @ WWW2013}{2013, Rio de Janeiro, Brazil}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

The current version of our framework is the result of a knowledge elicitation process performed internally in our lab. Our goal is to present it to the broader community and extend and revise it based on their feedback in order to achieve a shared understanding of the basic notions and of the associated terminology, and to consensually lay out the structure of the design space that will enable a systematic study of existing and future social machines. We anticipate that the framework will be a useful tool for both researchers in social and computer sciences, and for developers and operators of social machines. Using a common conceptualization allows the former to become familiar with the landscape, and identify topics of research that so far have remained unexplored or under-explored. They are given a theoretical grounding to observe the effects of specific technical properties of a system configuration on social behavior, discover design and evolution patterns of such systems, learn about social network formation and dynamics, and devise incentive mechanisms to encourage wide participation. For developers and operators, the framework may inform the engineering of new systems in terms of critical features and community development.

@@

The remaining sections are organized as follows. In order to pin down what makes up a social machine, and establish the boundaries of this emerging field of research, we need to first understand the relationship between social machines and related topics such as human computation, collective intelligence, and crowdsourcing; Section \ref{sec:comparison} gives a summary of this comparative analysis. In Section \ref{sec:framework} we then present the framework in terms of the methodology used to devise it, the main clusters of constructs used for classification, and an example of using them to compare a selection of popular social machines via grid analysis. We conclude with an outlook of our future development activities, including details about the evaluation of the framework. In addition, we plan to engage with the broader community to set up a `social machine' bringing together researchers from various disciplines who will contribute to the `computation' of a joint and widely agreed classification framework for the field.

@@

Wisdom of the crowds \cite{surowiecki2005wisdom} refers to a principle for decision making that takes into account information and opinions of a group of people rather than individuals; the use of specific technologies, most notably Web $2.0$, has made it possible for such processes to be carried out at scales hardly conceivable in the past, and to involve highly diverse and geographically distributed participants. A similar concept, though broader scoped, is collective intelligence, defined in \cite{malone2009harnessing} as `groups of individuals doing things collectively that seem intelligent.' The main difference to what we refer to as `social machines' is hidden in the the automation part of the latter. In a social machine, human and computational intelligence coalesce in order to achieve a given purpose. The overall setting may slightly vary in terms of the interaction patterns between the two types of services and the way their outcomes are used; however, where wisdom of the crowds and collective intelligence clearly place their focus on identifying the situations in which groups of people perform better than individuals, social machines are concerned with the study and realization of hybrid systems where the two types of components co-exist. As such, theoretical and empirical insights from the two related areas are useful to understand the dynamics of the social structures underlying such a system, but they are definitely not the only ingredient needed to build operational social machines. The question of how human and automatic services can be brought together to achieve optimal results, as well as the actual engineering exercise by which a system is developed, tested, and updated are equally important.

@@

An analogous line of reasoning could be followed to point out the overlaps between social machines and social computing \cite{parameswaran2007research}. The latter is an area of computer science which refers systems that support `the gathering, representation, processing, use, and dissemination of information that is distributed across social collectivities such as teams, communities, organizations, and markets' \cite{parameswaran2007research}. As such, compared to the general concept of `Computer-supported collaboration', social computing puts a greater emphasis on the information management capabilities of groups and communities, and less on the way these capabilities emerge as a joint effort. This distinction is even stronger in the case of social machines, which look at the social and the technical components as equal and necessary partners, and study the ways they could be best combined to master the challenges of future socio-technical systems.

@@

While determining the intersection of elements was straightforward, the consolidation of the
constructs required a more thorough process. We manually grouped the constructs into rough clusters, based around the areas they cover. We examined each construct to determine which were equivalent, and whether we could re-word or subsume existing constructs to cover the same aspects. The aim of this exercise was twofold: to remove any redundancy and cut the classification space down to a manageable size, while ensuring that all constructs that were elicited were represented in the final set. This process involved four of the authors discussing the choices, and resulted in a consolidated set of $31$ constructs organized in four different clusters: popularity, tasks and purpose, participants and roles, and motivation and incentives. In our analysis we will not further consider the first cluster, which primarily covered descriptive information about a given system, such as its perceived level of maturity and current number of users. Further on, the constructs were not evenly distributed across clusters; whereas some aspects such as roles of participants occurred frequently (in various flavors) in the grids provided by the ten experts, others such as the different types of workflows synthesizing human and automatized capabilities of a social machine were less taken into account. As discussed in Section \ref{sec:comparison} we believe that the interaction between these two types of services is an essential part of the theory and practice of social machines, being one of the elements that distinguishes them from related areas which are biased towards one or the other. Extending the framework with constructs capturing these aspects (see, for instance, \cite{quinn2011human} for examples in the context of human computation) is part of our future work.

@@

\subsection{The polyarchical relationship of social machines}
When defining the boundaries of what we call social machines, one important observation was related to the distinction between platforms and technologies such as wikis and GWAP (`games-with-a-purpose'), which enable social machines to be created, and their instantiations into social machines that were brought to life for a specific purpose, such as Wikipedia and EteRNA\footnote{\url{http://eterna.cmu.edu}}, a game in which participants design RNA that is evaluated automatically using simulators. A second observation is concerned with the relationship between broader and narrower-scoped instances of social machines; general-purpose examples such as Facebook, Twitter or Amazon, or even the Web, enable the formation of more specific social machines and communities within them, that are working to solve solve different problems. An example illustrating this relationship is the Obama'12 US Presidential campaign~\cite{obamakieron}, which relied on Twitter and Facebook as basic social machines; most forms of customer relationship management and community engagement carried out in the digital space use similar tools. Finally, social machines are interconnected into a greater ecosystem, both at the social and at the computation levels. For example, insights on the classification of space objects gained through GalaxyZoo may influence the content of the corresponding articles, and the associated editorial processes, within Wikipedia. In the same time,a large number of communities within Wikipedia focus specific topics and activities, such as those contributing to science and technology.\footnote{Visualization of Wikipedia Science Communities: \url{http://www.olihb.com/WikiCommunities/}}

Looking at the set of social machines in a polyarchy leads to a broad/specific relationship emerging that lets us talk about behavior at various levels of granularity. We propose looking at nested machines with `The Web' as one of several potential roots, with the next level down consisting of sub-platforms (such as Facebook, Twitter, MediaWiki, Ushahidi) that spawn more specific social machines. A resulting polyarchy is shown in Figure~\ref{polyarchy}, which classifies levels as infrastructure, frameworks, services, and projects and initiatives.

@@

The polyarchy is complementary to the classification constructs we elicited. The constructs apply to all levels of the polyarchy, though some of them might be more easier to understand and describe in the context of some levels. For example, instances of social machines residing at the bottom of the polyarchy are likely to be built with a very concrete audience in mind, and as such the mechanisms to incentivize participation are likely to be clearer and more straightforward to study and adjust than in cases in which such boundaries cannot be defined, such as Facebook, or even Wikipedia. In the same time, these systems will have to decide among an array of diverse basic services and frameworks for their implementations; the impact of such engineering decisions is difficult to predict in great detail as a thorough understanding of what makes certain social machines more successful than others, in terms of technology choices and beyond, is largely missing for most application domains.

%One such example is the `open access' movement in scholarly publishing
%~\cite{harnad2001self}. This cause mobilizes academic authors to self-archive their own papers online so they are available for free, in addition to hosted versions behind publishers pay-walls. In order to achieve this aim, the social machine of `open access' typically uses the social machines of institutional repositories provided
%by the authors' associated institutions. As with other social machines, these social
%machines typically use off-the-shelf implementations (in this case the most popular are
%EPrints~\cite{eprints} and DSpace~\cite{dspace}), which run as web sites on the world wide
%web. This cause is mature, growing year-on-year and has global coverage.

@@

\begin{table}[htb]
\begin{center}
\begin{scriptsize}
\begin{tabular}{|p{8cm}|}
\hline
{\bf Tasks, purpose and context of participation} \\
\hline
Activities involve creative production of content \\
Activities involve subjective appraisal of content \\
Activities involve solving (a definable) computation task or set of problems\\
Tasks are domain-specific \\
The machine owner derives benefit from participation \\
Activities and tasks are pre-defined or participant-defined \\
Variation in types of contributions and tasks \\
Participants' participate to fulfill needs of a role \\
\hspace{1cm} work-related/professional role \\
\hspace{1cm} home/family related \\
\hspace{1cm} social role \\
\hspace{1cm} leisure/entertainment role \\
Participation is done via: \\
\hspace{1cm} mobile devices \\
\hspace{1cm} Web browsers \\
\hspace{1cm} apps \\
\hspace{1cm} sensors/sensing (location sensing and wearable devices) \\
Participation is done in a mobile context \\
\hspace{1cm} Physical location is relevant to the service \\
\hline
{\bf Participants and roles} \\
\hline
Generality of audience \\
Participant autonomy \\
Participant anonymity \\
Extent of hierarchical organization of roles \\
Clear separation of roles among participants \\
{\bf Motivation and incentives} \\
\hline
Participants are intrinsically motivated:\\
\hspace{1cm} to gain/share knowledge \\
\hspace{1cm} to ``get something done'' \\
\hspace{1cm} to ``for fun/entertainment''\\
\hspace{1cm} to ``be social'' \\
\hspace{1cm} for the benefit of a specific group of people who need help \\
\hspace{1cm} for the benefit of society as a whole \\
Participation is motivated by extrinsic reward (payment, status) \\
\hline

@@

%% {\bf The types of tasks undertaken by the social component and the resulting contributions:}Social machines, just as any computing system, are built with a given purpose in mind, and exhibit functionality by which this purpose can be achieved. Users and the emerging social network interact with the system and with each other to carry out specific tasks by which they contribute to the `computation'. This includes the creation and refinement of digital artifacts by executing given actions, possibly in collaboration with other participants. For example, in GalaxyZoo\footnote{\url{http://www.galaxyzoo.org}} people classify galaxies with collective performance as good as professional astronomers, while in reCaptcha\footnote{\url{http://www.google.com/recaptcha}} they refine the data that describes digital manuscripts. Sometimes, systems look for innovative ways of getting people to engage with the systems, including game narratives, like in EteRNA\footnote{\url{http://eterna.cmu.edu}}, in which participants design RNA that is evaluated automatically using simulators and then refined via human effort, or monetary rewards, like in Amazon's Mechanical Turk, a platform in which users undertake routine (micro) tasks such as object classification, audio transcription and text translation.

@@

From the dendrograms we can get a sense of similarities and separation of social machines. For
instance, it is clear that YouTube, Vimeo, Reddit, and Digg are similar, which is not unexpected, due to their focus on sharing content. Likewise, the exercise confirms commonalities between Quora and StackOverflow, which can be easily understood, given their question answering purpose. There is also a dendrogram for the constructs, which can be used, as more data is collected, as an indicator for correlation between constructs. In this particular case, dendrogram suggests that there is a correlation between the `Single Participant Roles' and `No Hierarchy in Participant Roles' constructs, which is because they refer to the same aspect of social machines. Another, but less obvious connection is between the `Large variety of social features' and `Benefit NOT for the world/society', which requires further investigation.

>>>>

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{SOCM2013: Workshop on Theory and Practice of Social Machines, WWW2013}{2013, Rio de Janeiro, Brazil}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

The current version of our framework is the result of a knowledge elicitation process performed internally in our lab. Our goal is to present it to the broader community and extend and revise it based on their feedback in order to achieve a shared understanding of the basic notions and of the associated terminology, and to consensually lay out the structure of the design space that will enable a systematic study of existing and future social machines. We anticipate that the framework will be a useful tool for both researchers in social and computer sciences, and for developers and operators of social machines. Using a common conceptualisation allows the former to become familiar with the landscape, and identify topics of research that so far have remained unexplored or under-explored. They are given a theoretical grounding to observe the effects of specific technical properties of a system configuration on social behaviour, discover design and evolution patterns of such systems, learn about social network formation and dynamics, and devise incentive mechanisms to encourage wide participation. For developers and operators, the framework may inform the engineering of new systems in terms of critical features and community development.

@@

The remaining sections are organised as follows. In order to pin down what makes up a social machine, and establish the boundaries of this emerging field of research, we need to first understand the relationship between social machines and related topics such as human computation, collective intelligence, and crowdsourcing; Section \ref{sec:comparison} gives a summary of this comparative analysis. In Section \ref{sec:framework} we then present the framework in terms of the methodology used to devise it, the main clusters of constructs used for classification, and an example of using them to compare a selection of popular social machines via grid analysis. We conclude with an outlook of our future development activities, including details about the evaluation of the framework. In addition, we plan to engage with the broader community to set up a `social machine' bringing together researchers from various disciplines who will contribute to the `computation' of a joint and widely agreed classification framework for the field.

@@

Wisdom of the crowds \cite{surowiecki2005wisdom} refers to a principle for decision making that takes into account information and opinions of a group of people rather than individuals; the use of specific technologies, most notably Web $2.0$, has made it possible for such processes to be carried out at scales hardly conceivable in the past, and to involve highly diverse and geographically distributed participants. A similar concept, though broader scoped, is collective intelligence, defined in \cite{malone2009harnessing} as `groups of individuals doing things collectively that seem intelligent.' The main difference to what we refer to as `social machines' is hidden in the the automation part of the latter. In a social machine, human and computational intelligence coalesce in order to achieve a given purpose. The overall setting may slightly vary in terms of the interaction patterns between the two types of services and the way their outcomes are used; however, where wisdom of the crowds and collective intelligence clearly place their focus on identifying the situations in which groups of people perform better than individuals, social machines are concerned with the study and realisation of hybrid systems where the two types of components co-exist. As such, theoretical and empirical insights from the two related areas are useful to understand the dynamics of the social structures underlying such a system, but they are definitely not the only ingredient needed to build operational social machines. The question of how human and automatic services can be brought together to achieve optimal results, as well as the actual engineering exercise by which a system is developed, tested, and updated are equally important.

@@

An analogous line of reasoning could be followed to point out the overlaps between social machines and social computing \cite{parameswaran2007research}. The latter is an area of computer science which refers systems that support `the gathering, representation, processing, use, and dissemination of information that is distributed across social collectivities such as teams, communities, organisations, and markets' \cite{parameswaran2007research}. As such, compared to the general concept of `Computer-supported collaboration', social computing puts a greater emphasis on the information management capabilities of groups and communities, and less on the way these capabilities emerge as a joint effort. This distinction is even stronger in the case of social machines, which look at the social and the technical components as equal and necessary partners, and study the ways they could be best combined to master the challenges of future socio-technical systems.

@@

While determining the intersection of elements was straightforward, the consolidation of the
constructs required a more thorough process. We manually grouped the constructs into rough clusters, based around the areas they cover. We examined each construct to determine which were equivalent, and whether we could re-word or subsume existing constructs to cover the same aspects. The aim of this exercise was twofold: to remove any redundancy and cut the classification space down to a manageable size, while ensuring that all constructs that were elicited were represented in the final set. This process involved four of the authors discussing the choices, and resulted in a consolidated set of $31$ constructs organised in four different clusters: popularity, tasks and purpose, participants and roles, and motivation and incentives. In our analysis we will not further consider the first cluster, which primarily covered descriptive information about a given system, such as its perceived level of maturity and current number of users. Further on, the constructs were not evenly distributed across clusters; whereas some aspects such as roles of participants occurred frequently (in various flavours) in the grids provided by the ten experts, others such as the different types of workflows synthesising human and automatised capabilities of a social machine were less taken into account. As discussed in Section \ref{sec:comparison} we believe that the interaction between these two types of services is an essential part of the theory and practice of social machines, being one of the elements that distinguishes them from related areas which are biased towards one or the other. Extending the framework with constructs capturing these aspects (see, for instance, \cite{quinn2011human} for examples in the context of human computation) is part of our future work.

@@

\subsection{The polyarchical relationship of social machines}
When defining the boundaries of what we call social machines, one important observation was related to the distinction between platforms and technologies such as wikis and GWAP (`games-with-a-purpose'), which enable social machines to be created, and their instantiations into social machines that were brought to life for a specific purpose, such as Wikipedia and EteRNA\footnote{\url{http://eterna.cmu.edu}}, a game in which participants design RNA that is evaluated automatically using simulators. A second observation is concerned with the relationship between broader and narrower-scoped instances of social machines; general-purpose examples such as Facebook, Twitter or Amazon, or even the Web, enable the formation of more specific social machines and communities within them, that are working to solve solve different problems. An example illustrating this relationship is the Obama'12 US Presidential campaign~\cite{obamakieron}, which relied on Twitter and Facebook as basic social machines; most forms of customer relationship management and community engagement carried out in the digital space use similar tools. Finally, social machines are interconnected into a greater ecosystem, both at the social and at the computation levels. For example, insights on the classification of space objects gained through GalaxyZoo may influence the content of the corresponding articles, and the associated editorial processes, within Wikipedia. In the same time,a large number of communities within Wikipedia focus specific topics and activities, such as those contributing to science and technology.\footnote{Visualisation of Wikipedia Science Communities: \url{http://www.olihb.com/WikiCommunities/}}

Looking at the set of social machines in a polyarchy leads to a broad/specific relationship emerging that lets us talk about behaviour at various levels of granularity. We propose looking at nested machines with `The Web' as one of several potential roots, with the next level down consisting of sub-platforms (such as Facebook, Twitter, MediaWiki, Ushahidi) that spawn more specific social machines. A resulting polyarchy is shown in Figure~\ref{polyarchy}, which classifies levels as infrastructure, frameworks, services, and projects and initiatives.

@@

The polyarchy is complementary to the classification constructs we elicited. The constructs apply to all levels of the polyarchy, though some of them might be more easier to understand and describe in the context of some levels. For example, instances of social machines residing at the bottom of the polyarchy are likely to be built with a very concrete audience in mind, and as such the mechanisms to incentivise participation are likely to be clearer and more straightforward to study and adjust than in cases in which such boundaries cannot be defined, such as Facebook, or even Wikipedia. In the same time, these systems will have to decide among an array of diverse basic services and frameworks for their implementations; the impact of such engineering decisions is difficult to predict in great detail as a thorough understanding of what makes certain social machines more successful than others, in terms of technology choices and beyond, is largely missing for most application domains.

%One such example is the `open access' movement in scholarly publishing
%~\cite{harnad2001self}. This cause mobilises academic authors to self-archive their own papers online so they are available for free, in addition to hosted versions behind publishers pay-walls. In order to achieve this aim, the social machine of `open access' typically uses the social machines of institutional repositories provided
%by the authors' associated institutions. As with other social machines, these social
%machines typically use off-the-shelf implementations (in this case the most popular are
%EPrints~\cite{eprints} and DSpace~\cite{dspace}), which run as web sites on the world wide
%web. This cause is mature, growing year-on-year and has global coverage.

@@

\begin{table}[htb]
\begin{center}
\begin{scriptsize}
\begin{tabular}{|p{8cm}|}
\hline
{\bf Tasks, purpose and context of participation} \\
\hline
Activities involve creative production of content \\
Activities involve subjective appraisal of content \\
Activities involve solving (a definable) computation task or set of problems\\
Tasks are domain-specific \\
The machine owner derives benefit from participation \\
Activities and tasks are pre-defined or participant-defined \\
Variation in types of contributions and tasks \\
Participants' participate to fulfil needs of a role \\
\hspace{1cm} work-related/professional role \\
\hspace{1cm} home/family related \\
\hspace{1cm} social role \\
\hspace{1cm} leisure/entertainment role \\
Participation is done via: \\
\hspace{1cm} mobile devices \\
\hspace{1cm} Web browsers \\
\hspace{1cm} apps \\
\hspace{1cm} sensors/sensing (location sensing and wearable devices) \\
Participation is done in a mobile context \\
\hspace{1cm} Physical location is relevant to the service \\
\hline
{\bf Participants and roles} \\
\hline
Generality of audience \\
Participant autonomy \\
Participant anonymity \\
Extent of hierarchical organisation of roles \\
Clear separation of roles among participants \\
{\bf Motivation and incentives} \\
\hline
Participants are intrinsically motivated:\\
\hspace{1cm} to gain/share knowledge \\
\hspace{1cm} to ``get something done'' \\
\hspace{1cm} to ``for fun/entertainment''\\
\hspace{1cm} to ``be social'' \\
\hspace{1cm} for the benefit of a specific group of people who need help \\
\hspace{1cm} for the benefit of society as a whole \\
Participation is motivated by extrinsic reward (payment, status) \\
\hline

@@

%% {\bf The types of tasks undertaken by the social component and the resulting contributions:}Social machines, just as any computing system, are built with a given purpose in mind, and exhibit functionality by which this purpose can be achieved. Users and the emerging social network interact with the system and with each other to carry out specific tasks by which they contribute to the `computation'. This includes the creation and refinement of digital artefacts by executing given actions, possibly in collaboration with other participants. For example, in GalaxyZoo\footnote{\url{http://www.galaxyzoo.org}} people classify galaxies with collective performance as good as professional astronomers, while in reCaptcha\footnote{\url{http://www.google.com/recaptcha}} they refine the data that describes digital manuscripts. Sometimes, systems look for innovative ways of getting people to engage with the systems, including game narratives, like in EteRNA\footnote{\url{http://eterna.cmu.edu}}, in which participants design RNA that is evaluated automatically using simulators and then refined via human effort, or monetary rewards, like in Amazon's Mechanical Turk, a platform in which users undertake routine (micro) tasks such as object classification, audio transcription and text translation.

@@

From the dendrograms we can get a sense of similarities and separation of social machines. For
instance, it is clear that YouTube, Vimeo, Reddit, and Digg are similar, which is not unexpected, due to their focus on sharing content. Likewise, the exercise confirms commonalities between Quora and Stack Overflow, which can be easily understood, given their question answering purpose. There is also a dendrogram for the constructs, which can be used, as more data is collected, as an indicator for correlation between constructs. In this particular case, dendrogram suggests that there is a correlation between the `Single Participant Roles' and `No Hierarchy in Participant Roles' constructs, which is because they refer to the same aspect of social machines. Another, but less obvious connection is between the `Large variety of social features' and `Benefit NOT for the world/society', which requires further investigation.
